---
title: "Overview of PhyloProfileData"
author: "Hannah Mülbaier & Vinh Tran"
date: "`r Sys.Date()`"
package: PhyloProfileData
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{PhyloProfileData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction
The PhyloProfileData package contains the following data objects:

1. A preprocessed NCBI taxonomy data required by PhyloProfile package, and

2. Two experimental datasets to illustrate running and analysing phylogenetic 
profiles with PhyloProfile.

```{r, warning=FALSE, message=FALSE}
library(PhyloProfileData)
```

# Preprocessed NCBI taxonomy data
The *PhyloProfile* package provides two unique features in comparison to other
existing phylogenetic profiling tools. One of them the ability to rapidly change
the resolution of the comparative analysis of proteins in individual species, to
that acros classes, phyla or entire kingdoms by collapsing the input taxa into 
the corresponding systematic ranks. The other feature is that, *PhyloProfile* 
can automatically order the input taxa in increasing taxonomic distance from a 
user-specified reference taxon. Furthermore, *PhyloProfile* can convert the 
input taxon IDs into the scientific names for the presentation of the profiles,
which enable a more informative picture than showing only the taxon IDs. 

Those features are only available when we link the taxon IDs in the input 
phylogenetic profile with their full taxonomy information including the IDs and 
the scientific names from the lowest systematic rank (e.g. `strain` or 
`species`) up to the root of the species tree (e.g. `cellular organism`).

There are packages that can get the NCBI taxonomy ranks and IDs for a given 
taxon (such as *taxize*), however getting these information on the fly for a 
large number of taxa and parsing them can be time consuming. Therefore we 
provide a preprocessed taxonomy file for *PhyloProfile* that contains required 
information for those related functions. 

```{r, warning=FALSE, message=FALSE}
# Load data
data("ncbiTaxonNamesFull")
head(ncbiTaxonNamesFull)
```

The preprocessed taxonomy file contains the NCBI taxonomy ID for each taxon, 
their corresponding scientific name, the current systematic rank and the ID of 
its higher rank (parent ID). These information were parsed from the `names.dmp`
and `nodes.dmp` files downloaded from 
ftp://ftp.ncbi.nlm.nih.gov/pub/taxonomy/taxdmp.zip. This data contains 
2.110.551 entries, correspond to the complete NCBI taxonomy database (Federhen, 
2012) on May 13th 2019.

# Experimental datasets

## Phylogenetic profiles of AMPK-TOR pathway

The phylogenetic profiles of 147 human proteins in the AMPK-TOR pathway across 
83 species in the three domains of life were taken from the study of Roustan et 
al. 2016. 

This data set includes 3 files: 

- A phylogenetic profile input contains the human protein IDs, the taxaonomy ID
of the 83 search species and the corresponding orthologous protein IDs in those 
species, together with two additional values, the forward and backward FAS 
scores. FAS score is blablabla (please describe the FAS score here, with reference. 
This info you also need for your thesis and project report(Could't find any
paper or reference I could use to support my knowledge))

```{r, warning=FALSE, message=FALSE}
data("ampkTorPhyloProfile")
head(ampkTorPhyloProfile)
```

- A multiple fasta object contains the FASTA sequences for all the proteins 
present in the data set.

```{r, warning=FALSE, message=FALSE}
data("ampkTorFasta")
head(ampkTorFasta)
```

- A data frame contains the domain annotation for the proteins present in 
the phylogenetic profiles. The protein domain annotations were done using 
different annotation tools and databases, such as *PFAM*, *SMART*, *CAST*, 
*COILS2*, *SEG*, *SignalP*, and *TMHMM*. The annotation types together with 
their domain name and the corresponding start and end positions are stored in 
this domain data frame.

```{r, warning=FALSE, message=FALSE}
data("ampkTorDomain")
head(ampkTorDomain)
```


## Phylogenetic profiles of Arthropoda proteins

The phylogenetic profile of 1011 ortholog groups across 89 species 
was calculated from the BUSCO arthropoda dataset published
from Simão et al.2015. The BUSCO arthropoda dataset was downloaded from https://busco.ezlab.org/datasets/arthropoda_odb9.tar.gz in Jan. 2018. 

HaMStR-OneSeq, a tool which predicts orthologs and compares their 
Feature Architecture Similarity scores, bases the calculation on given 
core-ortholog datasets. To make the application more flexible and customized
due to the research question, the well known BUSCO datasets were transformed
to the format required from HaMStR-OneSeq, so they can be used as core-orthologs.

The BUSCO datasets mentioned above provided the information, which protein sequence
belongs to which BUSCO ortholog group, and was used to download the protein 
sequences from orthoDB version 10 (https://www.orthodb.org).
Afterwards, the sequences of every ortholog group were filtered off 10 species 
(Ladona fulva, Agrilus planipennis, Polypedilum vanderplanki, Daphnia magna, 
Harpegnathos saltator, Zootermopsis nevadensis, Halyomorpha halys, 
Heliconius melpomene, Stegodyphus mimosarum,Drosophila willistoni), the multiple
sequence alignments were computed with *MUSCLE* and the pHMMs were calculated 
with *HMMER3*. After the integration of these new groups as core-orthologs in 
HaMStR-OneSeq and the integration of the new species datasets to the one 
supplied by HaMStr, the phylogenetic profile was computed. 

This datset includes 3 files:

- A phylogenetic profile input contains the human protein IDs, the taxaonomy ID
of the 89 searched species and the corresponding orthologous protein IDs in those 
species, together with two additional values, the forward and backward FAS 
scores which is descriped above in the description of the AMPK-TOR pathway.

```{r, warning=FALSE, message=FALSE}
data("arthropodaPhyloProfile")
head(arthropodaPhyloProfile)
```


- A multiple fasta object contains the FASTA sequences for all the proteins 
present in the data set.

```{r, warning=FALSE, message=FALSE}
data("arthropodaFasta")
head(arthropodaFasta)
```

- A data frame contains the domain annotation for the proteins present in 
the phylogenetic profiles. The protein domain annotations were done using 
different annotation tools and databases, such as *PFAM*, *SMART*, *CAST*, 
*COILS2*, *SEG*, *SignalP*, and *TMHMM*. The annotation types together with 
their domain name and the corresponding start and end positions are stored in 
this domain data frame.

```{r, warning=FALSE, message=FALSE}
data("arthropodaDomain")
head(arthropodaDomain)
```

# References
1. Federhen, S. (2012) The NCBI Taxonomy database. Nucleic Acids Res, 40, 
D136–D143.
2. Roustan, V. et al. (2016) An evolutionary perspective of AMPK–TOR signaling 
in the three domains of life. Journal of Experimental Botany, 67 (13), 3897–3907
3. Simão, F. et al. (2015) BUSCO: assessing genome assembly and annotation 
completeness with single-copy orthologs. Bioinformatics published online June 9,
2015